Efficiency

### 2018-01-27 DraftExec.Rnw
couple of minor graphics/caption tweaks, centering images etc.
Run through the Rstudio spell checker for full document.  Now can't compile.  Wish I knew why.

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

The development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.

I should hasten to add that my expertise in this area is derived largely from an academic background whithin a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.  I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.

Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.  This paper doesn't seek to advance the technical field with any new research or insight.  It aims only to bring together as neatly and simply as possible those concepts and results which are neccessary and sufficient to give the appropriate weighting to this question.

@@

However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.  Scientific discovery and technical tends to confirm many of the best discoveries are actually built-upon and confirmed, rather than simply sweep them aside  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.  My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outining the foundational principals.

In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise embodiments in present design).  To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence. I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.

With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.  In addition, for increased transparency, the relevant code and equations underlying are available on github.

@@

The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.  Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines.

@@

The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.  The conclusion of this report is that the following statement can be made with confidence:
\\
\\
\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurace, and a lower overall range, than the one with a conventional propeller.  This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
\\
\\
Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.  One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.  This report finds no evidence of a pumpjet of the type used on submarine ever being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt. During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.  Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.  Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.  The basis for claims made by public officials to the contrary, such as those by Rear Admiral Sammut \parencite{sammut2017}, must be called into question.

This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.  These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthmore, a particular hydrodynamic phenonemon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.  At very low waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

@@

In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.  The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems  \parencite{giles2010}.  Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modelled in all directions, and the combinations combined and tested.

The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the interested parties.

@@

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

@@

The consequences of flow separation also has acoustic consequences, which this report explores.  The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.  This results in pressure fluctuations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.  These effects are are likely to cause some accoustic signature which becomes lower in frequency as the submarine slows down further and the vortexes become larger, and shed more slowly.  This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequencies as speed increases and bubbles expand.  Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.

@@

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction to movement.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

@@

\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}

@@

This has a profound impact on the operation and engineering of maritime vessels.  Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.  To the extent that time is non-critical, slower is always much much better.  It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming \parencite{liang2014} in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods \parencite[8-10]{sanguri2012}. The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.  Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.

@@

The same consideration applies to other types of vessels, which might use water-jets.  Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds. However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.  If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt. \footnote{The physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.}  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.

@@

Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load (total drag force at any given speed), but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.  Consequently, there are relevantly few pieces of literature which address the present question directly. However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.

@@

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.  It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal than diesel-electric submarines would, even at their peak output.

However, the difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage. Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

@@

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).  This keeps the total level of Xenon in check.

@@

\subsection{Recent Commentary}
Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet. However on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

@@

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.  One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably wait for a couple of days.

>>>>

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst the controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}. The pumpjet featured extremely prominently in the all the marketing and literature during the CEP \parenctite{costello2016}. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various discussion \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators remain reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone seems to acknowledge a trade-off exists, no-one seems confident enough to nominate the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the program to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was present, and was worthy of consideration.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.

As someone with some formal training in physics, and having considerable time considering military maritime technology running a small start-up defence company, I found this seemingly unresolved series of exchanges regarding the pumpjet choice particularly intriguing.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary in one respect or another. In the case of the pumpjets, which are claimed to be superior from an acoustic perspective when accelerating to higher speeds, the existence of some trade-off in propulsive efficiency (particularly at low speeds), appeared almost inevitable, potentially quite significant in scale.  Waterjets and pumpjets are widely held in maritime engineering circles to have their greatest advantage at very high speeds, but to be generally undesirable for low and medium speeds.  Confirming the validity of that view, and quantifying the scale of the impact, seemed to be an important and obvious 'common-sense' check for the credibility of the design concept put forward by the French Naval Group, and adopted by the Australian Government.

It is important also to acknowledge that assessing trade-offs requires more than a detailed understanding about one specific characteristic.  It's true that efficiency isn't everything in military technology, at least not in all circumstances.  In recent years, I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purposes, which generally operated at high speeds.  However, for some roles and operations (including survey or mine-countermeasures) other systems may have been more efficient. Balancing other desirable characteristics is an important and complex process.  In the case of submarines, the circumstances which lead to cavitation, and other drivers of acoustic performance are obviously closely related.  Consequently, the concurrent consideration of the link between propulsion systems and acoustic performance was a necessary feature of a paper attempting to shed light on this topic.  In addition, the propagation of efficiency comparisons through to a variety of particular metrics of direct importance to submarine operations and tactics (dived range, endurance, and indiscretion ratio) are also necessary and important to allow meaningful assessment of the trade-offs with other considerations to take place.

I am very glad to acknowledge that the development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the support of Gary Johnston.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest their own resources in research to advance public debate.  Unless he had commissioned me to write this report, it is unlikely that I would have been able to make the sustained and deliberate inquiry which the topic required to reach the scope and depth required to meaningfully inform debate.

I should hasten to add that my expertise in this area is derived largely from an academic background within a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.  I don't claim to be a global authority on this topic, but reasonably capable of grasping the works and words of those who are, which I have sought to do in my research at some length.

Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.  This paper doesn't seek to advance the technical field with any new research or insight.  It aims only to bring together as neatly and simply as possible those concepts and results which are necessary and sufficient to give the appropriate weighting to this question.

@@

However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.  Scientific discovery and technical progress tends to confirm and build upon many of the best previous discoveries and concepts, rather than simply sweeping them aside.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum, as well as many essential principals of mechanics and thermodynamics, are just as true, right, and relevant today as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.  My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outlining the foundational principals.

In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise embodiments in present design).  To that end, in this paper I've undertaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence. I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.

With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.  In addition, for increased transparency, the relevant code and equations underlying are available on GitHub.

@@

The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.  Unlike other modern technologies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have adopted pumpjet technology for their conventional submarines, which is currently only used by nuclear submarines \parencite(coorey2016).

@@

The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.  The conclusion of this report is that the following statement can be made with confidence:
\\
\\
\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance, and a lower overall range, than the one with a conventional propeller.  This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
\\
\\
Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.  One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.  This report finds no evidence of a pumpjet of the type used on submarine ever being recommended as superior for efficiency in general, and even closely related technologies for more commercial use not advocated below 25kt. During my research, I encountered one domain expert who was confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.  Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.  Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.  The basis for claims made by public officials to the contrary, such as those by Rear Admiral Sammut \parencite{sammut2017}, must be called into question.

This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principles of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.  These principles strongly support the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthermore, a particular hydrodynamic phenomenon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.  At very low water-speeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

@@

In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publicly available information relating to the likely technologies involved in the future submarine.  The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems  \parencite{giles2010}.  Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modeled in all directions, and the combinations combined and tested.

The model has been made available publicly in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the interested parties.

@@

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electrical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

@@

Flow separation also has acoustic consequences, which this report explores.  The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.  This results in pressure fluctuations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.  These effects are likely to cause some acoustic signature which becomes lower in frequency as the submarine slows down further and the vortexes become larger, and shed more slowly.  This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower frequencies as speed increases and bubbles expand.  Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.

@@

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction to movement.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an object experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficient ($C_D$) which reflects how aerodynamic or hydrodynamic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

@@

\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and submarines not particularly small, Reynolds numbers are likely to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}

@@

This has a profound impact on the operation and engineering of maritime vessels.  Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.  To the extent that time is non-critical, slower is always much much better.  It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming \parencite{liang2014} in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods \parencite[8-10]{sanguri2012}. The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.  Other incremental costs and inefficiencies from operating 'off-design' are frequently outweigh by such a dramatic reduction in overall power demand.

@@

The same consideration applies to other types of vessels, which might use water-jets.  Fast pilot-boats, for instance, might spend quite a considerable time manoeuvering in and out of harbour or alongside at very low water-speeds. However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.  If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt. \footnote{The physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.}  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about efficiency at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.

@@

Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load (total drag force at any given speed), but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truly unique engineering question, seldom discussed in commercial applications.  Consequently, there are relevantly few pieces of literature which address the present question directly. However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.

@@

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.  It is fair to say that in terms of maximum power output, nuclear submarines could have something like 5-10 times as much power at their disposal than diesel-electric submarines would, even at their peak output.

However, the difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage. Consequently, whereas a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

@@

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absorbs the neutrons to become a different isotope).  This keeps the total level of Xenon in check.

@@

\subsection{Recent Commentary}
Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particularly in the presence of the pumpjet. However on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

@@

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down voluntarily during operations.  One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably wait for a couple of days.

This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.  It might be presumed that surfacing and remaining surfaced for a couple of days is likely to constitute an end to combat operations for a submarine that was located in theatre, presumably through enemy action.  However, the implied alternative remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted. This fact is emphasised by the DCNS claim that hotel load exceeds propulsive power at patrol speeds.  As we can gather from other public estimates of submarine submerged endurance \parencite{buckingham2008submarine}, roughly a couple of days is a very significant fraction of the entire dived endurance of a diesel-electric submarine, perhaps 50\%.  It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of equivalently sized diesel-electric submarines, in addition to the reactor, simply for safety purposes.
